Week 5 - Chapter 7 pt.1 - Analysis

### 2024-10-02 week5.Rmd
indexupdate
week5firstpush

---
title: "Week 5"
author: "Verooo"
date: "2024-10-02"
output: html_document
---

@@

# Notes while discussing other peeps plots
Using RSiena for evolutin of network 
Wave, Ego alters 
RSciena: Ego, alter, wave 
Reorganize into an array for RSciena

@@

Or just remove all isolates lol 
```{r}

@@

>>>>

---
title: "Week 5 - Chapter 7 pt.1 - Analysis"
author: "Verooo"
date: "2024-10-02"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

@@

# Notes while discussing other peeps plots

Using RSiena for evolutin of network Wave, Ego alters RSciena: Ego,
alter, wave Reorganize into an array for RSciena

@@

Or just remove all isolates lol

## decision tree . order!

Something I really need a grasp of (bc very poor prioritization)!

## data were scraped at author level bceasier if you want to look at author "characteristics"

# Analysing data - Chapter 7 (SNASS)

## custom functions

```{r}
fpackage.check <- function(packages) {
    lapply(packages, FUN = function(x) {
        if (!require(x, character.only = TRUE)) {
            install.packages(x, dependencies = TRUE)
            library(x, character.only = TRUE)
        }
    })
}

fsave <- function(x, file = NULL, location = "./data/processed/") {
    ifelse(!dir.exists("data"), dir.create("data"), FALSE)
    ifelse(!dir.exists("data/processed"), dir.create("data/processed"), FALSE)
    if (is.null(file))
        file = deparse(substitute(x))
    datename <- substr(gsub("[:-]", "", Sys.time()), 1, 8)
    totalname <- paste(location, datename, file, ".rda", sep = "")
    save(x, file = totalname)  #need to fix if file is reloaded as input name, not as x. 
}

fload <- function(filename) {
    load(filename)
    get(ls()[ls() != "filename"])
}

fshowdf <- function(x, ...) {
    knitr::kable(x, digits = 2, "html", ...) %>%
        kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) %>%
        kableExtra::scroll_box(width = "100%", height = "300px")
}
```

## packages

RSiena RsienaTwoStep: this packages assesses the ministep assumption of
RSiena but is useful for this tutorial devtools: to load from github
igraph: plotting tools. For a tutorial on plotting networks, see 8

```{r}
packages = c("RSiena", "devtools", "igraph")
fpackage.check(packages)
devtools::install_github('JochemTolsma/RsienaTwoStep', build_vignettes=TRUE)
packages = c("RsienaTwoStep")
fpackage.check(packages)

@@

## RSiena as ABM

### Stochastic Actor-Oriented Models,aka. SAOM (see SNASS book)

### 7.5.1 Sample an ego

```{r}
ts_net1

# plot the network 
net1g <- graph_from_adjacency_matrix(ts_net1, mode = "directed")
coords <- layout_(net1g, nicely())  #let us keep the layout
par(mar = c(0.1, 0.1, 0.1, 0.1))
{
    plot.igraph(net1g, layout = coords)
    graphics::box()
}
```

randomly select/sample an agent.

```{r}

set.seed(24553253)
ego <- ts_select(net = ts_net1, steps = 1)  #in rsienatwostep two actors may make a change together but here not
ego
```

### possible networks after ministep

```{r}
options <- ts_alternatives_ministep(net = ts_net1, ego = ego)
# options
plots <- lapply(options, graph_from_adjacency_matrix, mode = "directed")
par(mar = c(0, 0, 0, 0) + 0.1)
par(mfrow = c(1, 2))

fplot <- function(x) {
    plot.igraph(x, layout = coords, margin = 0)
    graphics::box()
}

lapply(plots, fplot)
```

###7.5.3 Network statistics (see chapter 9 for more)

```{r}
# count the number of ties ego sends to alters
ts_degree(net = options[[1]], ego = ego)
# or for all options
lapply(options, ts_degree, ego = ego)

# count the number of reciprocated ties
lapply(options, ts_recip, ego = ego)
```

In the package RsienaTwoStep there are functions for the following
network statistics *s* :

**degree:** ts_degree()\
**reciprocity**: ts_recip()\
**outdegree activity**: ts_outAct()\
**indegree activity**: ts_inAct(**)**\
**outdegree popularity:** ts_outPop()\
**indegree popularity:** ts_inPop()\
**transitivity:** ts_transTrip()\
**mediated transitivity**: ts_transMedTrip()\
**transitive reciprocated triplets**: ts_transRecTrip()\
**number of three-cycles**: ts_cycle3()

### 7.5.4 Evaluation function

fi(x)=Σkβksik(x),

x= network

f = evaluation function (how we evaluate the nework)

i = the ego "i"

summing over k (k are diff. statistics (in our case degree and
reciprocity))

beta = value of the statistic

s = the specific statistic for ego i for network statistics k for our
network x

we are basically counting for each stats how many we have and
multipyling it by...

```{r}
option <- 4 # example of prior 4th option of the network of ego 4 
ts_degree(options[[option]], ego = ego) * -1 + ts_recip(options[[option]], ego = ego) * 1.5
```

Value for the network OF ego FOUR is 0.5 (coming from (-1\*1) + (1\*1.5)
= 0.5)

Or you could use the `ts_eval()`.

```{r}
eval <- sapply(options, FUN = ts_eval, ego = ego, statistics = list(ts_degree, ts_recip), parameters = c(-1,
    1.5))
eval
print("network with maximum evaluation score:")
which.max(eval)
```

calculate the evaluation of all possible networks (10) of ego 4 :

```{r}
eval <- sapply(options, FUN = ts_eval, ego = ego, statistics = list(ts_degree, ts_recip), parameters = c(-1,
    1.5))
eval
print("network with maximum evaluation score:")
which.max(eval)
```

higher evaluation score means ego 4 is more likely to have this kind of
network (option 4)

Translating the evaluation scores we just got into probabilities.

we know all probabilities will be between 0 and 1

and we know all the statistics (in our case the 10) will sum up to 1

1/e\^-0.5 "e to the power of -0.5" = 0.6065306 (idk how to write this in
R so I do not get an error that object e was not found :(

### 7.5.5 Choice function

We sum all the evaluation scores e to power of (0.6 etc.)

```{r}
# forcing ego to make a choice 
choice <- sample(1:length(eval), size = 1, prob = exp(eval)/sum(exp(eval)))
print("choice:")
choice
# print('network:') options[[choice]]
```

When you run it multiple times, it is different one every time. (this is
related to the see we set in the beginning. usually you would want to
set the seed in a way that multiple people all get the same number)
**This example should ideally be 10**.

--\> This is how to perform ONE MINI STEP

If we want to do it for next ego (e.g. ego ) in the network, that ego's
network depends on our prior ego's network (here it was ego/person 4)

Network changes of ego 2 depend on changes of ego 1. WHo is ego 1 and
who is ego 2 is random.

--\> Monte Carlo simulation

This is Agent-based modeling (if you do not use this for estimation).
Aka. setting rules and seeing if the output is like what we expected and
then trial and error to simulate the output as close as possible.

## 7.6 Simulation example

Let us now simulate how the network **could** evolve
given:^[18](https://snass.netlify.app/socionetsm#fn18)^

-   starting point at T1T1 is `ts_net1`

-   rate is set to 2

-   we as scientists think only network statistics degree and
    reciprocity are important

-   `RSiena::siena07` has determined the parameters for these statistics
    are -1 and 1.5 respectively

-   We adhere to the ministep assumption and hence set `p2step` to
    `c(1,0,0)`

```{r}
rate <- 2
degree <- -1
recip <- 1.5
ts_sims(nsims = 1, net = ts_net1, startvalues = c(rate, degree, recip), statistics = list(ts_degree,
    ts_recip), p2step = c(1, 0, 0), chain = FALSE)  #not that rate parameter is automatically included. 
```

This is one of the possible outputs from all the choices the egos made
in the networks in this simulation.

If we repeat this thousands of times, we can approximate a likely
network constellation.

## 7.7 Estimation logic

*How does the estimation work?*

Well, that is quite complicated. But it goes something like
this:^[19](https://snass.netlify.app/socionetsm#fn19)^

1.  Define model: researcher includes effects

2.  initial parameter values of effects (commonly ‘0’)

3.  simulate an outcome network based on these parameter values

4.  [compare]{.underline} the network statistics of the
    [simulated]{.underline} outcome network with the
    [observed]{.underline} outcome network at (i.e. the *target values*)

    -   based on the included effects. Thus the simulated network may
        contain 10 ties, but the observed network may contain 20 ties.
        Apparently, with the current parameter values we underestimate
        the density of the outcome network.

5.  tweak/update parameter values in *some smart way*

    (there is plenty of maths on this but in our case manually: our
    network density was too low, we looked at degree and reciprocity. We
    can increase the value of degree and try to approximate a value
    where changes in degree don't really make a change anymore. \
    Then we would change reciprocity parameter, in our case: increase
    it. We will have more reciprocity but also automatically increasing
    ties because reciprocity DEPENDS on ties\
    Thus, we need to find a way to do this automatically that can adjust
    both numbers at the same time.\
    We do not go about this ourselves though bc [as one may see] this is
    not very efficient)\
    \

6.  GOTO 3 (BREAK RULE: until parameter values cannot be improved
    anymore / or reached good fit)

7.  simulate a bunch of outcome networks with the obtained parameter
    values and compare the *expected values* of statistics of the
    simulated outcome networks with the *target values* of the actual
    observed outcome network.

    -   [Estimate a TARGET VALUE (outcome network). In an example we did
        it class: degree= 4, reciprocity 1]

    -   we can assess the fit

    -   estimate SE of the parameters

Let us suppose network `s501` developed into `s502`. For more
information on these networks see `?s501`.

To estimate this network with
`ts_estim`^[20](https://snass.netlify.app/socionetsm#fn20)^, we can do
the following:

```{r}
# we do not calculate SE for now.
# ans <- ts_estim(net1 = s501, net2 = s502, statistics = list(ts_degree, ts_recip), p2step = c(1, 0, 0), # used two different networks 
#    conv = 0.01, phase3 = TRUE, itef3 = 50, verbose = TRUE)

# looking at final result 
# ans
```

#output:

```         
#>            estim        SE 
#> rate    5.122247 1.1833275 
#> degree -2.258332 0.1371387 
#> recip   2.428552 0.2102846
```

## 7.8 Interpretation of parameters

### Network statistics

positive parameter = " i like"

positive degrees? people like degrees (higher degrees)

Sometimes it's easier to look at **Ratios**

Dividing two probabilities

when probability [statistics of one network]{.underline} is [*one more
than* another network,]{.underline} the answer is the exponential of the
parameter value in your model (the value we got for the network)

(given that alllllll other things are same between two things = ceterus
paribus condition)

ceterus paribusis not fulfilledif we look at degrees and reciprocity
htough, since reciprocity increasing also automatically makes the
degrees increase

if stat, for reciprocity is 2 and degree is 1, then they are of the same
unit (as opposed to other units like age, or years etc. )

#### in paper report:

descriptives, statistics

simple: higher scores = more positive

and recalculate in probability ratio: take exponent of parameter value \
"if netw. increases in this stats by one the probability ration is
this."

(you could have HUGE parameters but they might not be relevant at all
for the macro level change in networks)

Model of RSiena is modeling on EGO LEVEL only

-   statements like gender homphiliy is more important in my network
    than educational homophily

do smth similar

test with Rsiena:

-   for egos: Given a choice, gender homphily is more important than
    education...

-   more nuanced: GIVEN the network at T1, ...

MATCH YOUR HYPOTHESIS TO YOUR METHODS
